% {{{ LaTeX prelude
\documentclass[12pt]{article}
\usepackage{fontspec}
\setmainfont[Ligatures={Common,TeX}]{Baskerville}
\newfontfamily\g[Ligatures={Common,TeX}]{Times New Roman}
\usepackage{url}
\usepackage{parskip}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{titles}
\usepackage{enumitem}
\setlist{nosep}
% }}}

% {{{ LaTeX document
\begin{document}

% {{{ Title page
\begin{titlepage}
\title{Notes on Plato's \book{Euthyphro}}
\author{Peter Aronoff}
\date{April-May 2013}
\maketitle
\thispagestyle{empty}
\end{titlepage}
% }}}

% {{{ Characters and setting
\section{Characters and setting}

\subsection{Characters}

The dialogue is a conversation between Socrates and Euthyphro.  They run into
each other outside the office of the (so--called) king archon ({\g ἡ τοῦ
βασιλέως στοά}).  Both men are there because they have court cases before the
king archon.

Euthyphro appears to be the same person mentioned in \book{Charmides} 396a2--b1
and 396d2--e3.\footnote{\citet{burnet1924} 85 and \citet{bailly2003} 16.}  He
knows Socrates reasonably well, and he is very surprised to see Socrates at
court at all.  When Socrates explains that he's being charged with corruption
of the youth and religious crimes, Euthyphro assumes that Socrates'
\foreign{daimonion}\footnote{For a complete list of references to the
\foreign{daimonion}, see \citet{bailly2003} 32.} is the reason.  Perhaps partly
because of the \foreign{daimonion}, he sees Socrates as a kindred spirit:
Euthyphro views himself as a prophet and an expert on religious matters.  Other
Athenians, according to Euthyphro, view both Socrates and Euthyphro as unusual
because of their religious beliefs.  Other than the two Platonic dialogues, we
have no other evidence about Euthyphro.\footnote{\citet{bailly2003} 16.}

\subsection{Setting}

Because the king archon was responsible for various state religious functions,
he also oversaw trials that had to do with
religion.\footnote{\citet{burnet1924} 82--83.}  That's why both Socrates and
Euthyphro need to see him in particular.  It seems that Euthyphro has already
finished with the king archon since at the end of the dialogue he runs
off.\footnote{\citet{burnet1924} 82.}

The prosecutor\footnote{This word is somewhat misleading.  In ancient Athens,
a prosecutor was a private citizen, not someone who worked for the state.  Only
individuals could start a court case, whether civil or criminal.} and the
defendant met the king archon twice before the trial.  The first meeting was
very preliminary.  The prosecutor gave the defendent a written version of the
charge and paid a fee.\footnote {\citet{brickhouse2004} 7--8.}  The second
meeting with the king archon was the \foreign{anakrisis}.  The
\foreign{anakrisis} was more elaborate.  The defendant entered a formal plea,
and both sides answered questions about the case from the king archon.  At the
end of this meeting the king archon decided whether or not to send the case
forward to trial.\footnote {\citet{burnet1924} 82--83 and
\citet{brickhouse2004} 8.}

We can't say for sure which meeting Socrates will be having, but I'm inclined
to think it's the first. My argument is simple: the way he talks about Meletus
suggests absolutely minimal information about him.  But if they had already
done the first meeting, I think Socrates would have a (slightly) better sense
about him.  We could also make a parallel argument about the trial itself.
Socrates knows very, very little, and this is consistent with a first meeting.
Having said that, the first meeting with the king archon seems to have been
pretty short, so this isn't a conclusive argument.

% }}}

% {{{ Introduction (2--5d7)
\section{Introduction (2--5d7)}

% {{{ Socrates' case (2--3e6)
\subsection{Socrates' case (2--3e6)}

Euthyphro is shocked to see Socrates at court and asks who is prosecuting him
and what the trial is about.  (He assumes that Socrates is the defendant and
not the prosecutor.)  Socrates says that Meletus is the prosecutor, and they
spend a few minutes not knowing who he is.  Then Socrates tells Euthyphro the
charges: harming the young men via religious impiety.  Socrates first discusses
harming the youth.  He claims to find Meletus admirable because he is so
concerned for the youth of the city, as a politician should be.  However,
Socrates is openly confused by the charge that he is not irreligious.  He
describes that part of the charges as strange ({\g ἄτοπα} 3b1), but he does
clarify that the specifics of that part of the charge are (1) that he
introduces new gods and (2) that he does not believe
in\footnote{\citet{burnet1924} 15 (on 3b3) argues that {\g νομίζειν θεούς} does
\emph{not} mean \phrase{believe in} but rather \phrase{worship} and that ``it
refers primarily to religious `practice' ({\g τὰ νομιζόμενα})."  See also his
note on \book{Apology} 24c1.  I need to think more about this.} the gods of
Athens.

Euthyphro immediately assocates the first part of the irreligion charge with
Socrates' \foreign{daimonion}.  He expresses sympathy with Socrates, and says
that people always laugh at and mock him too, when he offers prophecy in public
meetings.  Socrates notes that there's a big difference between being laughed
at and a trial for your life.

% }}}

% {{{ Euthyphro's case (3e7--4e3)
\subsection{Euthyphro's case (3e7-4e3)}

Euthyphro seems somewhere between coy and proud of his
case.\footnote{\citet{geach1966} 369 takes Euthyphro to be embarrassed and
reluctant to admit who he is prosecuting.  Bailly on the other hand
\citeyearpar{bailly2003} 37 sees him as ``cavalier and cocky".
\citet{nehamas1975} 290 describes Euthyphro best as ``sensitive, touchy, and,
in a slightly perverse way, proud."}  He is prosecuting his father for the
murder of a hired hand.  The details are such that (1) there is room to doubt
whether the crime was really murder or simply manslaughter with callous
indifference\footnote{This terminology is mine and not necessarily historically
or legally accurate.} and (2) many Athenians would feel that Euthyphro was
wrong to pursue the case no matter what.  For such people, a son should not
prosecute his father.  This kind of disapproval leads Euthyphro to make the
assertion that opens him up to Socratic elenchus:

\begin{quote}

    {\g ἀνόσιον γὰρ εἶναι τὸ ὑιὸν πατρὶ φόνου ἐπεξιέναι· κακῶς εἰδότες,
    ὦ Σώκρατες, τὸ θεῖον ὡς ἔχει τοῦ ὁσίου τε πέρι καὶ τοῦ ἀνοσίου} (4d9-4e2).

    For [they think] that it is impious for a son to prosecute his father for
    murder.  But, Socrates, they understand poorly how the divine stands
    concerning piety and impiety.

\end{quote}

Euthyphro strongly implies here that \emph{he} does understand piety and
impiety. This is exactly the kind of confidence that Socrates looks for and
challenges.

% }}}

% {{{ Pre-definition methodology (4e4--5d7)
\subsection{Pre-definition methodology 4e4--5d7}

Socrates says that he should become Euthyphro's student in order to answer the
charges against him.  There is certainly \emph{some} irony here, though not
everyone would agree on how to interpret it.

Without getting too far into the irony wars, I'll just say this.  My take is
that the primary irony is a form of dramatic irony, and the source of the irony
is not so much Socrates as Plato.  That is, the author Plato is ironically
suggesting to the reader that Euthyphro the character is likely to become yet
another victim of Socrates: he begins with confident assertion, but we expect
him to end in \foreign{aporia}.  By focusing on Plato and his readers, rather
than on Socrates the character or Socrates the person, I think that we can
avoid the nightmare of trying to decide whether or not Socrates is
sincere.\footnote{In retrospect, this is a complete cop out.  I really should
think more about Socrates' tone, his irony and his sincerity.}

In any case, Socrates lays out the formal question he wants answered at 5c8ff.

\begin{quote}

    {\g νῦν οὖν πρὸς Διὸς λέγε μοι ὃ νυνδὴ σαφῶς εἰδέναι διισχθρίζου, ποῖόν τι
    τὸ εὐσεβὲς φῂς εἶναι καὶ τὸ ἀσεβὲς καὶ περὶ φόνου καὶ περὶ τῶν ἄλλων; ἢ οὐ
    ταὐτόν ἐστιν ἐν πάσῃ πράξει τὸ ὅσιον αὐτὸ αὑτῷ, καὶ τὸ ἀνόσιον αὖ τοῦ μὲν
    ὁσίου παντὸς ἐναντίον, αὐτὸ δὲ αὑτῷ ὅμοιον καὶ ἔχον μίαν τινὰ ἰδέαν κατὰ
    τὴν ἀνοσιότητα πᾶν ὅτιπερ ἂν μέλλῃ ἀνόσιον εἶναι;}

    Therefore now, by Zeus, tell me what you were just now confidently
    asserting that you knew clearly: What sort of thing do you say is the holy
    and the unholy, both concerning murder and concerning everything else? Or
    isn't the pious the same thing in every activity, itself to itself and the
    impious in turn opposite of every pious, but itself the same as itself and
    having some single form insofar as impiety, everything which will be
    impious?

\end{quote}

My translation is neither beautiful nor especially clear, but as far as that
goes, it closely resembles the Greek.  In this passage Socrates lays out
a number of conditions that he wants Euthyphro's definition to meet.

\begin{itemize}
    \item The definition should be universal.  However Euthyphro defines pious,
        that definition should cover both murder and everything else (5d1).
        Piety should be ``the same in every activity" (5d1).
    \item The two terms \word{piety} and \word{impiety} should be opposites
        (5d2).  This seems to suggest that they should \emph{exclude} one
        another, perhaps?
    \item Piety and impiety should ``have some single kind"\footnote{Many
        people get hot and bothered because Plato uses the word {\g ἰδέα}
        (\foreign{idea}), and this is one of his terms for \word{Form} in its
        technical sense.  Most scholars assume that this dialogue precedes the
        full theory of forms, so this use causes some anguish.  See
        \citet{burnet1924} 111.} (5d4) inasmuch as they are what they
        are.\footnote{I'm unsure whether this third condition is really
        a restatement of the first or a new point.}

\end{itemize}

Note that the universality that Socrates demands here is \emph{not} primarily
a matter of words.  He's not looking to legislate or to discover how we talk
about things.  We can be sure of that because in this very passage, he happily
uses two different sets of opposed terms for what he wants defined:
\foreign{eusebes} versus \foreign{asebes} and \foreign{hosion} versus
\foreign{anhosion}.

Although the confidence of Euthyphro's response may surprise us, it is
characteristic of similar moments in other dialogues.  Euthyphro immediately
answers ``Absolutely" ({\g Πάντως δήπου}, 5d6).  We might have expected him to
ask ``Huh?" or at least to request clarification from
Socrates.\footnote{\citet{nehamas1975} 303.}  However, Euthyphro's answer is
unsurprising in the light of other Platonic dialogues.  As John Burnet says:

\begin{quote}

    Euthyphro appears to be quite familiar with the terminology used by
    Socrates, and accepts it without complaint, confusion or shock.  That
    becomes all the more striking when we find him boggling later at much more
    elementary things.  Plato always represents the matter in this way.  No one
    ever hesitates for a moment when Socrates talks of {\g ἰδέαι} and {\g
    εἴδη}, and Socrates never finds it necessary to explain the
    terms.\footnote{\citet{burnet1924} 112.}

\end{quote}

Similarly, Alexander Nehamas writes:

\begin{quote}

    [I]t is notorious that Socrates' interlocutors (even Meno) are quite
    willing to answer affirmatively{\ldots}On this view, the existence of
    universals is taken for granted in the Socratic
    dialogues.\footnote{\citet{nehamas1975} 303--304.}

\end{quote}

Scholars disagree about many things here:

\begin{itemize}

    \item Why do the interlocutors agree so easily?

    \item What do they think they're agreeing to?

    \item If they understand what Socrates wants, why do they do so poorly at
        satisfying his demands?

\end{itemize}
% }}}

% }}}

% {{{ Definitions (5d8--15c10)
\section{Definitions (5d8--15c10)}

% {{{ First definition of piety (5d8--6a5)
\subsection{First definition of piety (5d8--6a5)}

Here is Euthyphro's first definition:

\begin{quote}

    {\g τὸ μὲν ὅσιόν ἐστιν ὅπερ ἐγὼ νῦν ποιῶ, τῷ ἀδικοῦντι ἢ περὶ φόνους ἢ περὶ
    ἱερῶν κλοπὰς ἤ τι ἄλλο τῶν τοιούτων ἐξαμαρτάνοντι ἐπεξιέναι, ἐάντε πατὴρ ὢν
    τυγχάνῃ ἐάντε μήτηρ ἐάντε ἄλλος ὁστισοῦν, τὸ δὲ μὴ ἐπεξιέναι ἀνόσιον}
    (5d8--5e1).

    Piety is exactly what I am doing now, to prosecute someone who commits
    a crime, whether murder or theft of temples or anything at all of this
    kind, whether the criminal happens to be father or mother or anyone at all;
    and not to prosecute is impiety.

\end{quote}

The idea seems to be that piety requires you to pursue \emph{anyone} who
commits religious crimes.

Unfortunately Euthyphro derails the conversation unintentionally right away.
He says that people give him a hard time for prosecuting his father, but they
contradict themselves because they're perfectly happy to hear stories about
Zeus and Chronos physically assaulting \emph{their} fathers.  Socrates wonders
if maybe this sort of thing is why he's in trouble: When he hears such stories,
he protests.  This line of argument will come back to haunt Euthyphro when
Socrates attacks his second definition.

% }}}

% {{{ Socrates rejects the first definition (6a6--6d10)
\subsection{Socrates rejects the first definition (6a6--6d10)}

Socrates puts aside, for the moment, his dislike of such stories in order to
get to his more basic complaint: prosecuting religious criminals may be pious,
but according to Euthyphro many other things are also pious.  What Socrates
wanted was the single pious: some one thing that we can use as a yardstick
whenever we need to judge whether anything else is pious or impious.
Euthyphro's first definition, by his own admission, cannot serve this purpose,
so Socrates rejects it.

% }}}

% {{{ Second definition of piety (6e11--7a1)
\subsection{Second definition of piety (6e11--7a1)}

Euthyphro's second definition is simpler and more general:

\begin{quote}

    {\g τὸ μὲν τοῖς θεοῖς προσφιλὲς ὅσιον, τὸ δὲ μὴ προσφιλὲς ἀνόσιον}
    (6e11--7a1).

    That which is pleasing to the gods is pious, and that which is not pleasing
    to the gods is impious.

\end{quote}
% }}}

% {{{ Socrates refutes the second definition (7a2--8b6)
\subsection{Socrates refutes the second definition (7a2--8b6)}

Socrates praises the form of Euthyphro's second definition, but he attacks its
content.  He argues as follows:

\begin{enumerate}

    \item The gods war against each other, they disagree and there is enmity
        between them.

    \item Among humans, only \emph{some} disagreements lead to this kind of
        fighting and enmity.  For example, if people disagree about a number or
        measurement, that doesn't lead to such a fight.  Those kinds of
        problems are solved easily enough by counting, measuring or weighing.
        But disagreements about justice and injustice, beauty and ugliness,
        good and bad~--- those disagreements lead to fighting and enmity.  They
        do so because in those cases people ``are unable to come to
        a sufficient agreement" (7d3--4).

    \item The gods quarrels and enmity results from these same sorts of
        disagreements.  Hence, the gods disagree about moral matters.

    \item What each god believes to be beautiful, good and just, that god
        loves.  They hate the opposites of such things.

    \item Therefore, \emph{the same things} are both hated and loved by the
        gods.

    \item Therefore, \emph{the same things} are both pious and impious.

\end{enumerate}


Socrates uses Euthyphro's earlier contention that the gods fight among
themselves against him.  In particular, Socrates comes back to the case of
Zeus, Chronos and Olympus and their relationship to Euthyphro's prosecution of
his father.  Socrates points out that Zeus may approve of what Euthyphro is
doing, as a son who had to punish his father, but that Chronos and Olympus
would not approve, as fathers who were punished by their sons.  Thus, we are
unable to judge solely by this yardstick whether or not Euthyphro's
controversial prosecution is pious or not.  Using divine approval as a test
gives contradictory results in this case.  Anything that can give contradictory
results isn't what Socrates asked for.  So, Euthyphro's second definition fails
as well.

% }}}

% {{{ Digression about the second definition (8b7--9b11)
\subsection{Digression about the second definition (8b7--9b11)}

Euthyphro tries at first to save part of his definition.  He argues that all
the gods agree about one thing at least: that whoever kills someone unjustly
should be punished.  Without agreeing or disagreeing, Socrates picks up
Euthyphro's claim and argues as follows:

\begin{enumerate}

    \item Every person agrees that wrongdoers deserve punishment, but they
        disagree about who the wrongdoers are.  In particular, they deny that
        they are wrongdoers.

    \item If the gods fight as Euthyphro says that they do, then they must be
        in the same situation.  They all agree that wrongdoers deserve
        punishment, but they fight over who the wrongdoers are.

    \item Therefore, if Euthyphro wants to maintain his argument on behalf of
        his prosecution of his father, he would need to show that \emph{every}
        god agrees that the day-laborer who died as a result of his father's
        neglect died unjustly ({\g ἀκίκως} 9a3), i.e. culpably.

\end{enumerate}

% }}}

% {{{ Second definition of piety reformulated (9c1--9e9)
\subsection{Second definition of piety reformulated (9c1--9e9)}

Socrates doesn't especially care about Euthyphro's specific case against his
father, so he relieves him of that burden.  As he says, even if Euthyphro
convinces him that all the gods agree about this one case, Socrates would only
have learned about \emph{one case} not about piety in general, as he wants.

At this point, Socrates takes over slightly, and he feeds Euthyphro his next
definition:

\begin{quote}

    {\g ὃ μὲν ἂν πάντες οἱ θεοὶ μισῶσιν ἀνόσιόν ἐστιν, ὃ δ᾽ἂν φιλῶσιν, ὅσιον,
    ὃ δ᾽ἂν οἱ μὲν φιλῶσιν οἱ δὲ μισῶσιν, οὐδέτερα ἢ ἀμφότερα} (9d2--3).

    Whatever all the gods hate is impious, and whatever they [all] love [is]
    pious; but whatever some love and some hate, this is neither or both.

\end{quote}

Socrates makes sure that Euthyphro accepts this new definition, and at 9e1--3
Euthyphro restates it himself.  Thereafter, it becomes Euthyphro's argument
once again.\footnote{Is there something worth saying here about ``Say what you
believe"?}

% }}}

% {{{ Socrates refutes all forms of the second definition (10a1--11b5)
\subsection{Socrates refutes all forms of the second definition (10a1--11b5)}

This next argument is probably the dialogue's most important legacy.  In this
refutation, Socrates presents what is commonly known as ``The Euthyphro
dilemma".  First, I'll lay out the dilemma.  Then I'll briefly discuss the
difficulties in interpreting Plato's text here.

Here is what Socrates says initially:

\begin{quote}

    {\g ἆρα τὸ ὅσιον, ὅτι ὅσιόν ἐστι, φιλεῖται ὑπὸ τῶν θεῶν, ἢ ὅτι φιλεῖται,
    ὅσιόν ἐστιν;} (12a2--3)

    Is the pious loved by the gods because it is pious, or is it pious because
    it it loved (by the gods)?

\end{quote}

This suggests the following dilemma:

\begin{enumerate}

    \item Either the gods love what is pious because it is pious or it is pious
        because they love it.

    \item If it is pious because they love it, then that seems arbitrary.  What
        if they loved murder?  Would murder then be pious?

    \item If they love it because it is pious, then we will probably need to
        care more about \emph{why} they love it than simply \emph{that} they
        love it.

    \item So, either the gods are arbitrary overlords or their judgments merely
        track what is independently pious.  Either way, we can't answer our
        deepest ethical questions simply by referring to what the gods love.

\end{enumerate}

This dilemma has had a great deal of historical importance inside religious
considerations of ethics and outside of religion.

Unfortunately, Socrates doesn't stop there.  Euthyphro doesn't understand his
initial formulation of the question.  In order to clarify things, Socrates goes
on a tangent about how we describe various things.  He \emph{seems} to be
contrasting either active versus passive or states versus processes.  Either
way, the language and the logic of his (so-called) clarification\footnote{As
\cite{cohen1971} 2 says, ``Socrates agrees to `speak more plainly' ({\g
σαφέστερον φράσαι}) and then produces the most baffling part of the argument."}
has only made readers scratch their heads.  I'm going to ignore that whole can
of worms.  For an excellent and detailed breakdown of what Socrates says and
what it means, see \citet{cohen1971}.

% }}}

% {{{ Aporetic interlude: Socrates' effect on beliefs (11b6--11e1)
\subsection{Aporetic interlude: Socrates' effect on beliefs (11b6--11e1)}

After Socrates refutes Euthyphro's adjusted second definition, he again asks
that Euthyphro say ``zealously" (προθύμως) what piety and impiety are
(11b4--5).  Euthyphro reacts this time with clear frustration and confusion:

\begin{quote}

    {\g Ἀλλ᾽, ὦ Σώκρατες, οὐκ ἔχω ἔγωγε ὅπως σοι εἴπω ὃ νοῶ· περιέρχεται γάρ
    πως ἡμῖν ἀεὶ ὃ ἂν προθώμεθα καὶ οὐκ ἐθέλει μένειν ὅπου ἂν ἱδρυσώμεθα αὐτό}
    (11b6--8).

    Well, Socrates, I don't have any way to say to you what I know. For
    whatever we put forth moves around somehow despite us, and it refuses to
    stay where we set it.

\end{quote}

The general tenor of this complaint is familiar from other dialogues.  Socrates
picks up on the idea of movement and says that if the moving arguments were
his, Euthyphro could joke that he was like his ancestor Daedalus.  This recalls
\book{Meno} 97d where a similar reference to Daedalus can be found.  The
specific idea is that Daedalus was able to create moving statues, and Socrates
can make arguments move~---even those of other people.  Under his influence,
Euthyphro's arguments don't ``stay where he places them" and he's unable to
make a stable definition of piety.  From Euthyphro's point of view, it's
\emph{Socrates} who is to blame for the failure of all the previous
definitions.

This aporetic interlude is important in another way: Some scholars argue that
Plato frequently puts arguments that he favors \emph{after} such
interludes.\footnote{\citet{mcpherran1992} 221.  See also his references in
footnote 14 (235--6).}  This would give readers extra reason to pay attention
to the arguments after the interlude.

% }}}

% {{{ Third definition of piety (11e1--12e9)
\section{Third definition of piety (11e1--12e9)}

Socrates accuses Euthyphro of being lazy at this point, but he also offers to
help and this may be a tacit acknowledgement of Euthyphro's frustration or
limits.

Socrates suggests that piety is a part (or kind?) of justice. Initially, his
wording confuses Euthyphro:

\begin{quote}

    {\g ΣΩ. ἰδὲ γὰρ εἰ οὐκ ἀναγκαῖόν σοι δοκεῖ δίκαιον εἶναι πᾶν τὸ ὅσιον.

    ΕΥΘ. Ἔμοιγε.

    ΣΩ. Ἆρ᾽οὖν καὶ πᾶν τὸ δίκαιον ὅσιον; ἢ τὸ μὲν ὅσιον πᾶν δίκαιον, τὸ δὲ
    δίκαιον οὐ πᾶν ὅσιον, ἀλλὰ τὸ μὲν αὐτοῦ ὅσιον, τὸ δὲ τι καὶ ἄλλο

    ΕΥΘ. Οὐχ ἕπομαι, ὦ Σώκρατες, τοῖς λεγομένοις} (11e4--12a3).

    SO. Consider whether it doesn't seem necessary to you for everything pious
    to be just.

    EUTH. [It seems so] to me.

    SO. Therefore is everything just also pious? Or is the pious all just, but
    the just is not all pious, but some of what is just is pious and some is
    something else.

    EUTH. I don't follow your argument, Socrates.

\end{quote}

After going through a few examples, Socrates gets his point across to
Euthyphro.\footnote{\citet{bailly2003} 90 writes the following about Socrates'
approach at this point:

\begin{quote}

    The method of defining a general area into which an object falls, then more
    and more precisely delineating the object is called ``division," and may be
    introduced into philosophy for the first time in the \book{Euthyphro} (the
    \book{Sophist} has extended instances and discussion of it).

\end{quote}

I'm not sure I buy this, but it's an interesting comparison.}  Socrates then
asks Euthyphro to specify what part of justice piety is, and Euthyphro answers:

\begin{quote}

    {\g Τοῦτο τοίνυν ἔμοιγε, ὦ Σώκρατες, τὸ μέρος τοῦ δικαίου εἶναι εὐσεβές τε
    καὶ ὅσιον, τὸ περὶ τὴν τῶν θεῶν θεραπείαν, τὸ δὲ περὶ τὴν τῶν ἀνθρώπων τὸ
    λοιπὸν εἶναι τοῦ δικαίου μέρος} (12e6--9).

    Therefore, Socrates, holiness and piety appear to me to be this part of
    justice: the part concerned with the service towards the gods, and the part
    concerned with service towards people is the remaining part of justice.

\end{quote}
% }}}

% {{{ Socrates refutes the third definition (12e10--14a10)
\subsection{Socrates refutes the third definition (12e10--14a10)}

Socrates praises the form of Euthyphro's latest answer, but he says that he
needs ``some small" clarification (13a1).\footnote{\citet{bailly2003} 96
compares \book{Protagoras} 329b.  See also Columbo's ``Just one more thing."}
The small thing concerns the word {\g θεραπεία}.\footnote{Reasonable
translations include \defn{service}, \defn{care}, \defn{caretaking}.  However,
it's not the easiest idea to convey directly in English with one or two words.
{\g θεραπεία} is a verbal noun derived from {\g θεραπεύω}.  The verb means
\defn{attend to}, \defn{serve}, \defn{care for}, \defn{provide for},
\defn{treat medically}.  The verb or noun might be used in a variety of
situations and across a variety of relationships.  So, for example, humans
\defn{attend to} gods, but gods \defn{provide for} humans.  Doctors \defn{care
for} or \defn{see to} patients, trainers \defn{care for} animals, and farmers
\defn{tend to} their crops~---all these ideas can be expressed by this one
word.}  Socrates expresses shock that Euthyphro uses this word, and he assumes
that Euthyphro can't mean the normal thing by it.  Socrates then takes
a characteristic turn.  He quickly lists three commonplace examples of {\g
θεραπεία}: care for horses, dogs and cattle.  He then asks Euthyphro if
θεραπεία for the gods is like these cases.  Euthyphro says that it is the same
kind of thing.

Socrates continues by specifying how he understands {\g θεραπεία} in the normal
cases.  In all such cases, the person performing the {\g θεραπεία} aims ``at
some good and benefit for the one being cared for ({\g τοῦ θεραπευομένου})"
(13b7--10).  Euthyphro agrees, after which Socrates asks him to explain what
benefit men do for gods when they exhibit piety.  Euthyphro naturally shrinks
from this implication, and the trap is sprung.

Euthyphro clarifies that the {\g θεραπεία} is not one where people
\emph{improve} the gods; it's more like slavery.  Socrates accepts this, using
the term {\g ὑπηρετική}.\footnote{This term also refers to \defn{service}, but
the connotation is clearly servile or from lesser to better.  Presumably,
Euthyphro thinks this will save his definition from seeming to be impious
itself.}  Socrates then effectively returns to his earlier line of argument,
though now without the embarrassing connotations.  He says that in the case of
medicine or shipbuilding there is a clear end product that the service aims at,
and Euthyphro agrees.  Socrates asks Euthyphro to tell him the parallel goal of
piety.


Euthyphro fumbles around, and Socrates does not hide his frustration.  First,
Euthyphro returns to vague answers, telling Socrates that the product of piety
are ``many fine things" (13e14).  Socrates agrees, but he explains generals and
farmers \emph{also} produce many fine things, and yet we can still say in
summary that they aim at victory and the production of crops.  What, Socrates
asks again, is the summary end product of piety?  This will lead us to the
fourth and final definition of piety.

% }}}

% {{{ Fourth definition of piety (14a11--14d3)
\subsection{Fourth definition of piety (14a11-14d3)}

At this point, Euthyphro repeats his earlier disclaimer that it is a difficult
thing to be specific, but he will say that if you know how to say and do things
pleasing to the gods, that is piety.  He adds that such piety saves both
individuals and cities.  Socrates objects to the length of Euthyphro's response
for the second time, and he also says that Euthyphro was just on the cusp of
teaching him what piety is, but that he ``turned away"
(14c1--2).\footnote{\citet{mcpherran1992} leans heavily on this passage when he
argues that the dialogue suggests a definition of piety that Socrates himself
approves of.}  Nevertheless, although Socrates is clearly not thrilled, he
explains that he must follow where the respondent leads (14c3-5).

Through a series of short questions and answers, Socrates clarifies Euthyphro's
definition as follows.  Sacrifice and prayer amount to asking the gods for
things and giving them things.  So, on Euthyphro's latest account, piety is
a matter of a service to the gods whereby we give them things (through
sacrifice) and ask them for things (through prayer).  Euthyphro emphatically
agrees with this clarification.

% }}}

% {{{ Socrates refutes the fourth definition (14d4--15c10)
\subsection{Socrates refutes the fourth definition (14d4--15c10)}

Socrates argues as follows:

\begin{enumerate}

    \item Assume that piety is service to the gods in asking them for things
        and giving them things.  Asking \emph{correctly} means asking for what
        we need.  Giving \emph{correctly} means giving what they need.
        Thus,\footnote{Socrates himself draws attention to an inference here,
        at 14e6 with {\g ἄρα}, though it isn't clear to me how he's entitled to
        such a `thus'.} piety turns out to be a kind of commercial interaction
        with the gods.\footnote{The Greek term he uses is {\g ἐμπορική}.
        Euthyphro is not thrilled with the connotations, but he accepts that
        word.}  Socrates' point is that according to Euthyphro's latest
        explanation, piety is a matter of giving the gods what they want in
        exchange for what we want.  Euthyphro probably takes himself to be
        agreeing to a series of commonplaces.  After all, what he describes is
        a standard \foreign{do ut des} view of
        religion.\footnote{\citet{bailly2003} 103--104.}

    \item But if this is so, then Euthyphro should be able to say how we
        benefit the gods through prayer.  Otherwise, our commerce with them
        seems to amount to this: We humans get everything good in our lives
        from the gods, but the gods get nothing from us in return.

\end{enumerate}

Euthyphro is unsurprisingly unhappy with the idea that we benefit the gods by
praying to them. In response, he returns to the idea that they gain honors,
prizes and gratification.  The word for gratification is {\g χάρις}, and this
leads Socrates back to the idea that piety is what is pleasing ({\g
κεχαρισμένον}) to the gods.  Socrates asks, ``So is it then pleasing but not
beneficial or beloved?" (15b1--2).  Euthyphro answers that he certainly thinks
piety \emph{is} beloved by the gods, and at that point Socrates has him caught
again.

Socrates reminds Euthyphro that this was his earlier definition, and he
presents him with a dilemma: Either Euthyphro is wrong now, or he was wrong
earlier to abandon that definition when it first came up.  Euthyphro grudgingly
concedes that he is caught in this dilemma, but he doesn't come down on either
side of the matter.

% }}}

% }}}

% {{{ Conclusion (15c11--16a4)
\section{Conclusion (15c11--16a4)}

At this point Socrates proposes that they start all over again.  He says that
he remains determined to learn about piety from Euthyphro.  Euthyphro, however,
begs off, saying that he has an appointment.  Socrates is exceedingly and
verbally disappointed. The dialogue ends on that note.

% }}}

% {{{ Poscript
\section{Poscript}

% {{{ What's the point?
\subsection{What's the point?}

We should wonder: What is the point of this dialogue?  The answer is far from
obvious, at least to me.  One suggestion is that the dialogue offers a hidden
answer to the question \phrase{What is piety?}.  I'm not very persuaded by
this, but even if I were, I would then want to ask why Plato chooses to give an
answer in such a roundabout way.  On the other hand, if there's no answer in
the dialogue, then what is the point?

Are we supposed to learn, with Euthyphro and Socrates, that we don't know
something?  Does Euthyphro really learn this?  Or does he just leave in
frustration?  Are we supposed to learn something about Euthyphro and perhaps
other people like him?  If so, what is the lesson?  It can't be simply ``some
people think they know things, but really they don't."  Or at least, if it's
only that, then this dialogue (and the many like it) seem to be killing a fly
with a tank.  \emph{That} lesson could have been stated much more easily and in
far fewer pages.  Or might Plato disagree?  Does he maybe think that we are
that stubborn or blind to the ignorance of ourselves and the people around us?

A related question is whether we should read this dialogue alone or together
with other early dialogues.  Someone could argue, I think, that this dialogue
by itself simply frustrates, but that larger meanings come forth if we make
connections with other early Socratic dialogues.  On the other hand, this again
should make us wonder why Plato works as he does.

I'm drawn to the idea that Plato deliberately makes this more difficult for his
readers.  I don't yet have any very detailed explanation of why or to what
(ultimate) end, but I'm persuaded by this argument:

\begin{enumerate}

    \item The dialogues are difficult to interpret.  And not just a little bit
        difficult, but a lot difficult.

    \item Plato is not a poor writer or a weak thinker.

    \item Therefore, the difficulties we meet must be deliberate and not the
        result of bad writing, poor planning or stupidity on Plato's part.

\end{enumerate}

I grant that this still leaves most everything to be worked out.

% }}} What's the point?

% {{{ Socratic irony
\subsection{Socratic irony}

Let's at least try to state \emph{why} it's so hard to give simple answers
about Socratic irony.  As an example statement for analysis, I'll use what
Socrates says to Euthyphro early in the dialogue:

\begin{quote}

    {\g Ἆρ᾽οὖν μοι, ὦ θαυμάσιε Εὐθύφρων, κράτιστόν ἐστι μαθητῇ σῷ γενέσθαι}
    (5a3--4)

    Therefore, wondrous Euthyphro, is it best for me to become your student?

\end{quote}

Recall the context: Euthyphro has just said that he knows about divine matters
with great precision (4e4--5a2).  So Socrates proposes that he become
Euthyphro's student.  How are we to take this?

\begin{enumerate}

    \item Completely straight: Socrates simply states his desire to learn from
        Euthyprho.  No matter what we may think about Euthyphro's knowledge,
        Socrates is being sincere. He wants to know more about divine matters
        in order (1) to be a better person and (2) to win his case against
        Meletus. Therefore, he hopes to learn from Euthyphro.

    \item Completely ironic: Socrates thinks Euthyphro is an idiot, and he is
        mocking him, somewhat openly in fact.  It is only \emph{because}
        Euthyphro is such a dolt that he does not see what is clear to every
        reader of the dialogue: Socrates is about to cross-question him into
        \foreign{aporia}.

    \item Some mixture of irony and seriousness: In theory, we should strive to
        be nuanced. Let's admit some of the palpable irony without entirely
        dismissing the serious purpose behind Socrates' words.  However, I'm
        unsure how to do this without fudging or hand-waving.  When Socrates
        speaks that sentence above, how are we to parse it so that it can be,
        all at the same time, both ironic and straight?  I confess that I would
        like to find a ``middle ground", but I'm not sure that such a place
        would be stable and solid.

    \item Straight from Socrates, but ironic from Plato: I suggested something
        like this above. The idea is that we can have our cake and eat it too,
        without being wishy-washy or hand-wavy.  The sentence I quoted above is
        both ironic and straightfoward, but only in the sense that the speaker
        Socrates, as a character, is sincere while the author, Plato, is
        ironic.  I like this because it seems to preserve some of what is good
        about the previous idea without falling into an impressionistic
        appearance of depth without substance.

        This interpretation can also, perhaps, give us two things that we might
        want.  First, we get a Socrates (as a character) who is not a complete
        asshole.  If we put too much irony onto Socrates, he begins to look
        cruel and aggressive.  Perhaps he has this side to him, but it makes me
        nervous.  Second, this reading allows us to admit the very real sense
        that we feel as readers that \phrase{No~---that can't be serious. He
        can't \emph{really} mean that.}  Vlastos and many of his students give
        us a completely sincere Socrates, but they ignore the tone of many of
        the dialogues. I'm not convinced that this is a reasonable trade, all
        things considered.

\end{enumerate}

% }}} Socratic irony

% }}} Poscript

\newpage
\bibliographystyle{apa}
\bibliography{plato}

\end{document}
% }}}
