% {{{ LaTeX prelude
\documentclass[11pt]{article}
\usepackage{fontspec}
\setmainfont[Ligatures={Common,TeX}]{Baskerville}
\newfontfamily\g[Ligatures={Common,TeX}]{Times New Roman}
\usepackage{enumitem}
\setlist{noitemsep}
\usepackage{url}
\usepackage{parskip}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{titles}
% }}}

% {{{ LaTeX document
\begin{document}

% {{{ Title page
\begin{titlepage}
\title{Notes on Plato's \book{Laches}}
\author{Peter Aronoff}
\date{June 2013}
\maketitle
\thispagestyle{empty}
\end{titlepage}
% }}}

% {{{ Characters and setting
\section{Characters and setting}

% {{{ Characters
\subsection{Characters}

The dialogue has a relatively large number of speakers.  In addition to
Socrates, there are two fathers, Lysimachus and Melesias and two generals,
Laches and Nicias.

The fathers first.  Lysimachus and Melesias both had famous politicians for
fathers, Aristides the Just and Thucydides respectively.  But neither
Lysimachus or Melesias has amounted to much.  They blame their fathers for
this, believing that their fathers did not raise them properly, and they
would like to do better for their sons.  (Oddly, \book{Meno} (94a and 94c)
indicates that both men received an excellent education.  So it's not
entirely clear what their complaint is or whether it's a valid one.)
Lysimachus belongs to the same \foreign{deme} as Socrates, and it turns out
that he was good friends with Socrates' father as well.  According to
Thucydides the historian (VIII.86.9), Melesias was a member of the
oligarchy put in charge of Athens in 411.

The generals next.  Nicias was an important general and politican in the
time before and during the Peloponessian War.  His policies during the
Sicilian expedition, where he lost his life in 413, helped to cause the
Athenian's terrible defeat there.  Laches was also an important general
during the Peloponessian war.  He died in 418 at the battle of Manitinea.
Laches and Socrates fought side by side during the retreat at Delium.  When
Alcibiades describes Socrates behavior at Delium (\book{Symposium}
221a--b), he specifies that Socrates was ``more prudent" ({\g ἔμφρων}) than
Laches, but I'm not sure how much that evidence is worth.

% }}} Characters

% {{{ Setting
\subsection{Setting}

Lysis and Melesias have invited Laches and Nicias to view a display by
a teacher of hoplite fighting.  (The teacher's name is Stesilaus.  He
doesn't speak, and we know of him only from this dialogue.)  The dialogue
doesn't provide details about exactly where the men are, but clearly it is
a \foreign{gymnasium} of some kind.  Socrates just happens to be around, so
we might guess that the display takes place at one of his favorite haunts.
But obviously this is speculative---and probably not important since Plato
didn't make a point about it.

% }}} Setting

% }}} Characters and setting

% {{{ Introduction (178a--190d6)
\section{Introduction (178a--190d6)}

The introductory section of \book{Laches} is relatively long and involved.
Before we get to anything explicitly Socratic, we have the following.
Lysimachus makes a long speech inviting Nicias and Laches to help him and
Melesias decide whether to train their sons in hoplite fighting.  Nicias
and Laches agree, but Laches urges Lysimachus to involve Socrates as
well.\footnote{The fathers invited Nicias and Laches.  Socrates luckily
happens to be around.  His presence isn't very surprising though since
he spent much of his life in and around training grounds.}  Lysimachus
agrees, and he remembers that he and Socrates have an old family
connection: Lysimachus was friendly with Socrates's father.  Nicias
makes a speech in favor of training the boys in hoplite fighting, and
Laches makes a speech against such training.  Lysimachus urges Socrates
to break the tie, and only then do we reach something explicitly
Socratic as Socrates argues that expertise and not numbers should
decide the matter.  Socrates then moves the discussion in familiar ways
until the generals are giving definitions of \word{bravery}.

Although much of this material is not explicitly Socratic, a number of
familiar Socratic themes appear implicitly in the introduction.  First,
Lysimachus and Melesias are very concerned for the proper education of
their sons.  This is already Socratic.  But in addition they worry because
their fathers were excellent men and well-renowned politicians, but they
have not done so well.  In \book{Protagoras} Socrates argues that political
excellence is not teachable, partly on the grounds that the most esteemed
politicans were unable to pass on their success to their children
(319e--320b).  Nicias also makes a claim that hints at Socratic
intellectualism.  He argues that someone who has knowledge of hoplite
fighting will fare better in battle (182b2--4).\footnote{Initially this may
sound like simply a truism, but since the dialogue will eventually turn
on whether or not bravery is a form of knowledge, I think it's an
important bit of foreshadowing.}  Laches on the other hand argues that many
trainers think and pretend that they have skill as fighters, but they they
are shams.  This criticism implicitly recalls Socratic arguments showing
that people think they know things that they actually do not.  Laches
explicitly concludes by discussing people who ``think that they know"
(184b4) something and the great ``slanders" (184b7) and ``envy" (184c1)
that they face as a result.

Socrates becomes more actively involved at 184d5, and he guides things in
a familiar direction.  Lysimachus, as I mentioned above, wants Socrates to
break the tie between Nicias and Laches.  Socrates, however, objects on
principle to making the decision by counting votes.  He immediately gets
Melesias to agree that they must follow knowledge and not number (184d8-9).
Additionally he suggests that they don't even know what subject they need
an expert in yet.  This suprises Melesias and Nicias, but Socrates makes
a familiar argument:

\begin{enumerate}

    \item When someone seeks advice about a drug for their eyes, the
        decision is really about their eyes, not the drug.  And when
        someone seeks advice about a bit for a horse, the decision is about
        the horse not the bit.  Nicias agrees to both of these (185c5--d4).

    \item Therefore, more generally, when someone seeks advice about X for
        Y, the decision concerns Y and not truly X.  Nicias agrees to this
        too (185d5--8).\footnote{Unwisely, I think.}

    \item Therefore it is necessary to consider whether a fellow decision
        maker is an expert in the real goal of the discussion, not any
        proximate goal.  Again, Nicias agrees (185d9--12).

\end{enumerate}

Socrates then proposes that their consideration of training in hoplite
fighting is ``for the sake of the soul of the young men" (185e1--2).
Therefore they must find an expert in training young people's souls not
hoplite fighting.

Socrates assumes that if you an expert, then you should be able to prove
it.  His basic demands for proof are either (1) the name of your teacher
and proof of his \foreign{bona fides} or (2) evidence that you have taught
others well, if you have not had a teacher but discovered things on your
own.  Initially, it appears that Socrates will question Nicias and Laches
about their status as experts, but the dialogue finds a clever way to avoid
this.  Once they reach the point where it would be natural for Socrates to
test them on these matters, Socrates suddenly thinks of an even better way
to investigate: If they are really experts, then they should be able to
give an account of the thing they are experts in.  So instead of asking
them questions about their teachers or students, Socrates will simply ask
them about the subject in question.

The subject in question is virtue ({\g ἀρετή}), and they immediately
whittle that down to bravery.  Socrates worries that all of virtue would be
too large a topic, so he proposes that they discuss a part of virtue
(190c8-10).  It's important to notice that he is assuming here that virtue
has parts.  This assumption will play an important role later in the
dialogue.  Bravery is the obvious choice to pick, Socrates says, since they
began all this by talking about hoplite fighting.

% }}} Introduction (178a--190d6)

% {{{ Definitions of bravery (190d7--199e12)
\section{Definitions of bravery (190d7--199e12)}

% {{{ First definition of bravery (190d7--192b8)
\subsection{First definition of bravery (190d7--192b8)}

The first definition is a famous example of an interlocutor not giving
Socrates the right type of answer.  Laches defines bravery as ``being
willing to fight off the enemy while remaining in formation and not
fleeing" (190e5--6).  Importantly, Socrates doesn't deny that acting as
Laches describes is brave.  In fact, he explicitly agrees that it is
(191a1--3).\footnote{See \citet{nehamas1975} for why this passage is
\emph{not} an example of someone giving a particular when Socrates
wants a universal.  The answer Laches gives is not a particular, though it
is too specific.}  Nevertheless, Socrates is not satisfied with this
answer.

Socrates objects that Laches' definition does not cover \emph{all} brave
actions.

\begin{enumerate}

    \item The Scythians and some Homeric heroes flee while fighting, but
        they are brave (191a8--10)

    \item The Spartans at Plataea fled, but they were brave (191b8--c5)

    \item Bravery doesn't concern only one type of fighting.  It isn't even
        limited to fighting.  People also exhibit bravery while sailing and
        in the face of poverty, illness and politics.  Even more, people
        can exhibit bravery in the face of good things too: desires or
        pleasures (191c8--e2).\footnote{Laches agrees even to these last
        examples.  I'm not sure that he should have though.  Yes, someone
        can make good decisions concerning pleasure and desire, but are
        such cases really examples of bravery?}

\end{enumerate}

Laches agrees to try again, but he still isn't 100\% clear about what
Socrates wants in an example, so Socrates gives an example definition of
his own for speed.  Socrates first argues that speed, like bravery, occurs
in a variety of cases: running, playing an instrument, speaking, learning
and many others.  He then explains that his definition must fit all of
these cases and any others where speed occurs.  His definition is ``speed
is a power accomplishing many things in little time" (192a10--b2).  Laches
approves of the answer, and Socrates invites him to give a similar answer
for bravery.  This brings us to our second definition.

% }}} First definition of bravery (190d7--192b8)

% {{{ Second definition of bravery (192b9--194c6)
\subsection{Second definition of bravery (192b9--194c6)}

Laches shows that he paid attention to Socrates' criticism in his second
definition.  He defines bravery as ``a kind of endurance of the soul" ({\g
καρτερία τις τῆς ψυχῆς}, 192b8).  He is clearly trying to maintain the
spirit of his first definition while at the same time to satisfy Socrates'
demand for generality.  It would seem at first glance that \word{endurance}
is a broad enough idea to cover military cases, which are foremost in
Laches' mind, as well as other examples of bravery, such as those that
Socrates mentioned.  In addition, the addition ``of soul" may hint at the
later explorations of more intellectualist definitions by Nicias.

Socrates criticizes this definition for being too inclusive.  This is the
opposite of the first definition's problem, and so these two definitions
thus serve as an object lesson in Socratic definition: He wants all but
only F for his answer to ``What is F?".

Socrates makes his case that endurance is too broad in familiar ways.
First, Socrates introduces cases where the proposed \foreign{definiens}
would not be fine ({\g καλόν}) or good ({\g ἀγαθόν}) as all virtues should
be. Second, Socrates brings forward cases where the proposed
\foreign{definiens} is insufficiently grand to count as a virtue.  In cases
like this, Socrates relies on his interlocutors' shame at certain
associations.\footnote{I should get examples of other cases like this,
right?  In \book{Gorgias}, I know he uses a shameful case against either
Callicles or Polus---the shit-bird.}

\begin{enumerate}

    \item First, Laches agrees that bravery is fine (192c5--6).

    \item Then Laches agrees that thoughtful endurance is fine and good.

    \item But Laches also agrees that thoughtless endurance is harmful and
        evil-doing (192d1--3).

    \item Since this kind of endurance is not fine and bravery is fine,
        Laches concedes that this kind of endurance is not brave.  Only
        wise endurance can be brave (192d4--12).

    \item But Socrates restricts wise endurance as well.  Socrates asks
        Laches whether wise endurance applied to spending money would count
        as bravery?  What about a doctor who wisely endures the pleading of
        a sick patient for something harmful?  In both cases, Laches denies
        that these are examples of bravery (192e--193a2).\footnote{I wonder
        whether Laches would have been smarter to deny that these are
        examples of \word{endurance}?}

\end{enumerate}

The first part of this argument has intuitive appeal to me.  I would
redescribe it this way.  For any action type you name, some examples may be
good and some examples may be bad.  To use another Socratic example (from
\book{Republic} i), giving someone back the axe that he lent me is not
necessarily a good idea.  If the lender is drunk and enraged when he asks
for it back, perhaps I should hold onto it instead.  I would like to put
this point as follows: Whatever action we name should be done \emph{with
reasonable discrimination}.  The discrimination of circumstances is how
I would understand Socrates' demand for intelligence and thoughtfulness
here.

Socrates returns to military examples, but now forces a different objection
on Laches.  What about cases where a person endures wisely and is willing
to fight, but he is aware that he has a significant advantage over his
opponent(s)?  Intuitively Laches agrees that the opponent who is at
a disadvantage seems \emph{braver} than the one with the great advantages.
But that seems to run against the argument that wisdom and bravery go
together since Socrates claims (and Laches agrees) that the person who is
at a disadvange displays ``more foolish endurance" (193b2--3).

This argument troubles me.  Socrates seems to assume that wisdom tracks or
guarantees success, but this strikes me as wrong.  To begin with, it
assumes (or implies or relies on or flirts with?) the Socratic notion that
the good man cannot be harmed.  To the degree that we find that doubtful,
we might also doubt that failure implies a lack of intelligence or virtue.
Secondly, I wonder if Socrates takes too thin a view of the example he
considers.  Perhaps the people who are outnumbered and certain to die are
brave but not thoughtless.  They know that they are unlikely to win, but
this does not turn them away.  As an example, consider the Greeks at
Thermopylae.

However, Socrates immediately gives two other examples that may serve his
argument better.  First he asks whether someone who is experienced in
cavalry fighting is less brave than a novice if both enter a fight on
horseback; he makes a similar case for slingers or archers (193b5--10).
Then he considers divers.  If someone inexperienced in deep-well diving is
willing to dive to the bottom of the well, is that person braver than an
experienced diver (193c2--8)?  Laches is inclined to say the less
experienced person in each case is braver.

All of these examples allow Socrates to draw out a contradiction from
Laches' agreements so far:

\begin{enumerate}

    \item Unwise daring and endurance was agreed to be shameful and harmful
        (193d1--3).

    \item But bravery was agreed to be fine (193d4--5).

    \item But now the foolish endurance is said to be bravery (193d6--8).

\end{enumerate}

Socrates does not explicitly demonstrate the contradiction, but Laches
readily agrees that he doesn't like where the argument has led them.  He
expresses frustration and says that he finds himself unable to say what he
thinks he understands well.  Socrates is sympathetic, but says that they
must ``be brave" and ``endure", and he suggests that they invite Nicias to
help.  This will bring us to our third definition.

% }}} Second definition of bravery (192b9--194c6)

% {{{ Third definition of bravery (194c7--199e12)
\subsection{Third definition of bravery (194c7--199e12)}

Nicias offers the third and final definition of bravery: bravery is a kind
of wisdom (194d8--9).  In response to questions by Socrates, Nicias
clarifies that wisdom is ``knowledge of fearful and bold things"
(194e11--195a1).  It's important that Nicias claims to have heard these
ideas from Socrates.  Presumably then, we should pay special attention to
them.  Socrates acknowledges that he has said such things, but he never
says explicitly whether he favors Nicias' theory himself.  In addition, he
ultimately disproves it---at least in Nicias' opinion---and so we should be
careful not to say that what we see in \book{Laches} is Socrates' theory.

% }}} Third definition of bravery (194c7--199e12)

% {{{ Objections from Laches (195a2--196c6)
\subsubsection{Objections from Laches (195a2--196c6)}

Laches strenuously and rudely objects to Nicias.  In fact, Laches is
genuinely unpleasant in the remainder of the dialogue.  Socrates asks him
more than once to try to take Nicias' ideas seriously and not simply insult
the man, and Nicias himself says (fairly) that Laches seems more interested
in Nicias being wrong than in discovering the truth.  My sense is that all
of this helps to show how infuriating Socratic ideas and arguments could be
to ordinary people, even when they didn't come from Socrates.

Laches attempts to use examples of people who know the right things but are
not brave.  His first example is doctors: They know what is fearful, but
(as Nicias agrees) they are not brave (195b3--6).  His next example is
farmers, and Laches quickly generalizes to say that all skilled
craftspeople know what is fearful or brave in their respective areas, but
nevertheless none of them is brave (195b7--c1).

Nicias responds by arguing that doctors know about health or illness, but
they don't know about what is (really?) fearful.  He unpacks this by saying
that they don't know whether it is fearful for someone to be healthy or to
be sick.  As he goes on to argue (and Laches agrees), some people are
better off dead.  A doctor cannot know which people these are: his
knowledge is limited to health and illness, but he cannot say whether you
are the sort of person who should live and be healthy (195c7--d9).

Laches objection to this is that Nicias sets the bar too high: his brave
people must be prophets (195e1--4).  Nicias however denies that even
a prophet has the requisite knowledge.  Again Nicias argues that a prophet
might know what is going to happen but not whether what is going to happen
is or is not good for someone (195d8--196a2).

Laches becomes so frustrated at this point that he turns the argument over
to Socrates.  He continues to snipe at Nicias as Socrates speaks to him,
but he doesn't contribute positively to the debate after this point.  He is
frustrated because Nicias seems to him to have made it impossible for
anyone but perhaps ``some god" to meet the standards for bravery (196a6).
In addition, he feels that Nicias is arguing eristically and changing his
stance and position merely to avoid being caught out.

Laches simply cannot take Nicias' proposals seriously, and I think that
this is no accident nor simply an indictment of Laches' temper in debate.
The point, I think, is that this is how people react to the wildly
counter-intuitive proposals of Socrates.  (Again, remember that Nicias has
taken his definition from Socrates.)  The anger, disbelief and frustration
result from the violent clash of common sense and Socratic argument.  What
Nicias proposes flies in the face of much that Laches (and many other
Athenians) took for granted.  It requires a significant amount of patience
and openness even to consider it seriously.

% }}} Objections from Laches (195a2--196c6)

% {{{ Objections from Socrates (196c7--e12)
\subsubsection{Objections from Socrates (196c7--e12)}

Socrates objects first that according to Nicias' definition, animals would
not be brave (196e2--7).  This runs counter to the Greek commonplace that
a soldier might be ``brave as a lion" (cf. \book{Iliad} \foreign{passim}).
Socrates tries to make the problem even more pointed by saying that lion
and stag and bull and monkey would all be equally brave (192e7--9).
(Presumably, one would object that a lion is obviously braver than a stag
and so forth.)

Nicias, however, accepts this consequence without blinking.  He agrees that
animals cannot be brave, and he adds that there is no bravery when there is
fearlessness as a result of ignorance.  As Nicias puts it ``lack of fear
and bravery are not the same thing" (197b2).  Nicias also explicitly
acknowledges here that his use of the words \word{bravery} and \word{brave}
run counter to what ``many people" (197b4) say.  This too does not bother
him.

Socrates next objection, and the argument that ends the dialogue
ultimately, is that Nicias contradicts himself in the following way.

\begin{enumerate}

    \item Nicias agrees that bravery is a part of virtue; other parts of
        virtue are temperance, justice and more (197e10--198b1).

    \item Nicias agrees that fear and courage have to do with the future.
        One feels fear over something bad that one believes is going to
        happen.  One feels courage in cases where one believes that nothing
        bad is going to happen.  Nicias also agrees that according to him
        bravery is the knowledge of these matters (198b2--198c8).

    \item Changing tack, Socrates gets Nicias to agree that domains of
        knowledge, the knowledge is not split between times.  That is, if
        I know about medicine, my knowledge covers health as it was, as it
        is, and as it will be.  We don't find three different areas of
        knowledge here: just one across all three possible times
        (198c9--199a9).

    \item But since bravery is a kind of knowledge, it too must be about
        not merely future events, but current and past ones as well.
        Therefore, they must emend their definition of bravery to include
        present and past (199a10--c2).

    \item If we say this, then bravery will include knowledge of all good
        and evil, not merely that which is going to happen (199c3--199d3).

    \item Therefore, bravery will be all of virtue, not merely a part of
        virtue (199d4--199e5).

    \item But they already agreed bravery was only a part of virtue.
        Therefore, they haven't found bravery (199e6--12).

\end{enumerate}

Now as many scholars\footnote{Get references here, please.} have said,
Nicias agrees to the final step but he need not have.  He could have
affirmed some version of the unity of the virtues in an effort to save the
argument.  Arguably \book{Laches} is one of those Socratic dialogues that
only apparently ends in \foreign{aporia}, but that has a strongly-implied
solution.

% }}} Objections from Socrates (196c7--e12)

% }}} Definitions of bravery (190d7--199e12)

% {{{ Conclusion (199e13--201c5)
\section{Conclusion (199e13--201c5)}

After Laches and Nicias bicker further, Laches repeats his earlier advice
to Lysimachus and Melesias: Don't let go of Socrates.  Nicias agrees, and
he too encourages the fathers to seek Socrates' advice and help.
Lysimachus asks Socrates, but Socrates begs off, saying that it makes no
sense to seek his advice since he was as clueless as the other two men.

As a counter-proposal, Socrates suggests that all of the adults, himself
included, seek a teacher for themselves.  Once they find a good teacher,
that person can also work with the young men.  But in Socrates' view, they
all need help, even though they are older.  He invokes Homer to support the
view that if a person is in need, shame is no help.  That is, they should
not be ashamed to admit that they need help and don't know what virtue or
bravery are.  Lysimachus agrees, and they agree to begin their search
tomorrow.  The dialogue ends with Socrates promising to visit Lysimachus
the next day for this purpose.

% }}} Conclusion (199e13--201c5)


\newpage
\bibliographystyle{apa}
\bibliography{plato}

\end{document}
% }}}
