% {{{ LaTeX prelude
\documentclass[11pt]{article}
\usepackage{fontspec}
\setmainfont[Ligatures={Common,TeX}]{Baskerville}
\newfontfamily\g[Ligatures={Common,TeX}]{Times New Roman}
\usepackage{enumitem}
\setlist{noitemsep}
\usepackage{url}
\usepackage{parskip}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{titles}
% }}}

% {{{ LaTeX document
\begin{document}

% {{{ Title page
\begin{titlepage}
\title{Notes on Plato's \book{Lysis}}
\author{Peter Aronoff}
\date{June 2013}
\maketitle
\end{titlepage}
% }}}

% {{{ Characters and setting
\section{Characters and setting}

% {{{ Characters
\subsection{Characters}

Socrates speaks with a number of young men in this dialogue.\footnote{The
information below is taken from \citet{race1983} and \citet{watt1987}.}

\begin{enumerate}

    \item Hippothales is a teenager and the lover of Lysis.  At the start
        of the dialogue, he spots Socrates and calls him over.  So, they
        clearly have some relationship.  He does not appear elsewhere in
        Plato, but Diogenes Laertius says that he was a follower of
        Socrates (III.46).

    \item Ctesippus is around Hippothales' age and a friend of his.  He
        mocks Hippothales for being so madly in love with Lysis.  He is
        Menexenus cousin, and both of them were present at the death of
        Socrates in \book{Phaedo} (59b).

    \item Menexenus is the same age as Lysis, and both of them are younger
        than Hippothales or Ctessipus.  He is Ctessipus cousin and
        a follower of Socrates.  He is probably the same Menexenus as
        appears in the dialogue of that name.

    \item Lysis is a close friend of Menexenus and the beloved of
        Ctesippus.  We don't know of him other than what is said in this
        dialogue.

\end{enumerate}
% }}} Characters

% {{{ Setting
\subsection{Setting}

The conversation takes place near and inside a newly opened gymnasium run
by Miccus.  Hippothales says that Miccus is an admirer and companion of
Socrates, but we don't know anything about him other than what's said in
this dialogue.  The gymnasium is somewhere between the Academy and the
Lyceum since Socrates says that he was coming from the former and heading
for the latter.  It's near a ``little gate by the spring of Panops"
(203a2--3).  \citet{race1983} places this ``near the Lyceum" and ``just
outside the eastern city wall."

% }}} Setting

% }}} Characters and setting

% {{{ Introduction (203a--207d4)
\section{Introduction (203a--207d4)}

Socrates offers advice on love to one Hippothales who is desperately in
lust with Lysis.  After Ctesippus, Hippothales and friends catch sight of
Socrates, Ctesippus reveals to Socrates that Hippothales is crazy about
Lysis.  What happens next is a little surprising: Socrates begins to offer
advice.  He tells Hippothales that he's a fool if he praises someone before
the person is caught.  (Praise just makes them arrogant, don't you know.)
Hippothales asks Socrates for help, and Socrates agrees to give it.  They
come up with a sitcom-worthy plan for Socrates to talk to Lysis within
earshot of Hippothales.

After starting a conversation with Lysis and Menexenus, Socrates turns the
conversation towards friendship ({\g φιλία}).  This will be the real topic
of the dialogue.  As \citet{watt1987} points out, there is an implicit
contrast at the start of the dialogue between {\g ἔρως} and {\g φιλία}, two
types of ``love" (119--120).  Hippothales lusts after Lysis, but Lysis and
Menexenus have deep friendship for one another.

% }}} Introduction (203a--207d4)

% {{{ The importance of wisdom (207d5--210d8)
\section{The importance of wisdom (207d5--210d8)}

Socrates leads Lysis through a complex question and answer session about
wisdom and friendship.  The upshot of the discussion seems to be that in
areas where you are wise, people will trust you and you will be free to do
as you please.  Indeed they will even turn their affairs over to you.  But
where you are not wise, you are a slave to others.  I wouldn't really call
this section an argument since so much of it relies on unlikely assertions
that receive no support.  It's also not an elenchus since Lysis doesn't
assert anything initially.

It's difficult to know how seriously to take the passage.  On the one hand,
many of the ideas seem Socratic.  On the other hand, at the end of the
discussion Socrates indicates that the whole conversation was an exercise
in humbling and demeaning an {\g ἐρόμενος} in an effort to woo him
(210e2--5).\footnote{Yes, that's right.  Socrates sounds like a modern pick
up artist game-player here.}  If we take that seriously, then I'm not sure
any of the preceding conversation should be taken very seriously.

Nevertheless here's how the conversation goes:

\begin{enumerate}

    \item Socrates gets Lysis to agree that his parents love him very much
        and that therefore they want him to be as {\g εὐδαίμων} as possible
        (207d5--8).

    \item Next Socrates suggests that a person is unhappy as a slave and
        unable to do what he or she wishes (207e1--3).

    \item Therefore, Socrates continues, if his parents love Lysis and want
        him to be as happy as possible, they must allow him to do as he
        wishes, not rebuke him and not prevent him from doing what he wants
        (207e3--7).

    \item But, says Lysis, they \emph{do} prevent him from very many
        things.  Socrates is shocked, just shocked (207e8-208a1).

    \item Socrates next asks about a number of specific things that Lysis
        might want to do.  In each case, Lysis is not allowed to do what he
        wants, but a paid servant or slave is allowed to do it.  For
        example, Lysis cannot take out one of his father's chariots
        whenever he wants and drive it as he likes.  In fact, Lysis cannot
        even fully control himself: He has a slave who takes him to school
        and the like.  Socrates dwells for a bit on the idea that a free
        person might in these ways be bettered by a slave even.  Lysis'
        father has assigned many people to rule over the boy, and his
        mother too does not let him do what he wants when it comes to her
        domain (280a1--209a4).

    \item Lysis says that they don't let him do these things because he is
        too young, but Socrates isn't having it.  In fact, his parents let
        Lysis do certain things regardless of his age.  They let him read
        and write for example.  And they let him handle instruments however
        he likes (209a4--209c1).

    \item These examples finally lead Lysis to draw a more Socratic
        conclusion: they don't let him act as he likes where he lacks
        knowledge; where he has knowledge, he can do as he wants (209c2).

    \item Socrates expands this beyond the family of Lysis.  A neighbor,
        the people of Athens, even the Great King would entrust things to
        Lysis, if they believed that he was knolwedgeable about those
        matters (209c6--210a8).\footnote{At this point I'm really unsure
        what's happening.  Up until now, things were borderline plausible.
        But I don't see why we (or Lysis) should swallow such huge
        assertions without argument.  Also, Socrates seems to wink in this
        passage: the Great King will entrust his \emph{meat-boiling} to
        Lysis.}

    \item Socrates sums up as follows.  In domains where we are
        knowledgeable, people will turn matters over to us.  In those
        cases, we will be free to do as we want. In addition we will own
        those things---they will be our own---insofar as we will profit
        from them.\footnote{The word \word{{g οἰκεῖος}} appears here
        (210c2), arguably foreshadowing its importance later in the
        dialogue. See also 210d2.}  But where we are without knowledge, we
        will not be in control, we will be at the command of other people,
        things will not belong to us insofar as we will not profit from
        them (210a9--c5).

    \item From all of this, Socrates draws some shocking conclusions.
        First we are not friends with someone nor is anyone friends with us
        unless there is benefit.  Second, and drawing on the first, parents
        do not love their children if the children are of no benefit to
        them.  Therefore, if you want friends and you want your parents to
        love you, you must become wise (210c5--210d4).

    \item As a final weird point, Socrates gets Lysis to agree that you
        can't be arrogant about areas where you're not wise (210c4--210c8).

\end{enumerate}

% }}} The importance of wisdom (207d5--210d8)

% {{{ Transition to Menexenus (210e1--211d5)
\section{Transition to Menexenus (210e1--211d5)}

As this first conversation between Socrates and Lysis winds down, the focus
turns to Menexenus.  He returns from helping with the sacrifice, and Lysis
is very eager for Socrates to speak with him as he spoke with Lysis
himself.  It seems that Lysis wants to humble Menexenus as well.\footnote
{See esepcially 211c3 where Lysis says that he wants Socrates to speak with
Menexenus "in order to punish him."}

Ctesippus also gets involved, and there's a funny sort of multi-way
flirting going on.  Ctesippus doesn't want Socrates and Lysis conversing
all to themselves, Hippothales is listening to all of this carefully, and
Lysis wants Socrates to beat up on Menexenus for him.  There's also a lot
of the kind of playful ``Do this", ``Well if you ask, then I must" sort of
thing that you see in flirting.

% }}} Transition to Menexenus (210e1--211d5)

% {{{ Socrates questions Menexenus about friends (211d6--213d5)
\section{Socrates questions Menexenus about friends (211d6--213d5)}

This section is a more traditional Socratic elenchus.  Socrates presents
Menexenus as a kind of expert, though Menexenus does not make such a claim
for himself.  Nevertheless, Socrates says that he has always prized
friendship very highly, and that he marvels at Lysis and Menexenus. For
they are so young, but have already acquired each other---such an excellent
friend each has.  Initially Socrates says that what is doesn't know is
``How does someone become a friend of someone's" (212a5--6), but this isn't
actually the question he pursues with Menexenus.

The question that Socrates actually asks Menexenus is odder: When one
person is friendly to another, who becomes whose friend? (212a8--b1).  As
Socrates goes on to explain, he wonders when one person befriends another
which of the following is the case (212a8--b2):

\begin{enumerate}

    \item The (active) befriender [becomes friend of] the (passive)
        befriended.

    \item The (passive) befriended [becomes friend of] the (active)
        befriender.

    \item No difference.

\end{enumerate}

Menexenus chooses the ``no difference" option.  It's not immediately clear
what ``no difference" means, but Socrates reply to Menexenus revels that
``no difference" amounts to ``both become friends of the other."  Socrates
immediately sets out to cross-question Menexenus on his choice.  His
argument is as follows.

\begin{enumerate}

    \item From Menexenus: If one person befriends another, then both become
        friends of each other (212b2--3).

    \item Sometimes someone befriends another person but is not befriended
        in turn by that person.  Menexenus agrees (212b5--6).

    \item Sometimes someone befriends another person, but that person hates
        the one befriending him.  Again, Menexenus agrees (212b5--c3).

    \item Therefore in cases like this, one person befriends and the other
        is befriended. [But it doesn't seem right to say that both become
        friends of the other in cases like this.] (212c3--4)

\end{enumerate}

Thus, Socrates asks again who is whose friend. And he offers three options,
though now the options are slightly different:

\begin{enumerate}

    \item The befriending person is a friend of the befriended one, whether
        he is befriended in return or even hated (212c5--6).

    \item The befriended person is a friend of the befriending one
        (212c6--7).

    \item Neither is a friend of the other unless both befriend each other
        (212c7--8).

\end{enumerate}

Menexenus goes for the new option: neither is a friend of the other unless
both are friends.  Socrates rephrases this as ``Nothing is a friend to
something befriending it unless it befriends in return" (212d4--5).  He
then proceeds to disprove this claim as follows:

\begin{enumerate}

    \item Assuming the given definition, nobody could be a lover of horses,
        a lover of quail, a lover of dogs, a lover of wine, of wrestling or
        of wisdom.

    \item For all such things cannot love in return.

\end{enumerate}

Menexenus rejects the conclusion, so they turn to the option that it is the
beloved that is a friend to the one befriending, whether the beloved loves
in return or not---even if the beloved hates the one befriending.  Socrates
offers an example of infant children.  They cannot love in return and even
sometimes seem to hate, but nevertheless they are most beloved to their
parents.

Although the first example seems fine, things get uglier when Socrates
generalizes.  In the generalized version, Socrates says that the befriended
is a friend, no matter the feelings of that person. And the hated is an
enemy, again regardless of the feelings of that person.  But this means
that many are befriended by their enemies and hated by their friends; and
it means that many are friends to their enemies and enemies to their
friends.  The absurdity of these results leads Socrates and Menexenus to
reject this claim as well.

At this point, Socrates regroups.  He reminds Menexenus that they have
rejected the following options\footnote{Did they actually disprove all
these options?  I'm not sure that they did.  Certainly they didn't handle
them in this order.}:

\begin{enumerate}

    \item The befriender is a friend [to whomever he befriends].

    \item The befriended is a friend [to whomever befriends him].

    \item Both befriender and befriended are friends [to each other].

\end{enumerate}

Socrates asks if Menexenus can think of any other people who can become
friends, since none of these worked out.  Menexenus, unsurprisingly, has no
ideas.  Socrates wonders whether they have not been going about the whole
thing wrong, and at this point, Lysis gets involved again.  Lysis says that
he does think they are going about things all wrong, and Socrates turns
back to him.

% }}} Socrates questions Menexenus about friends (211d6--213d5)

% {{{ Like is friend to like (213d6--215c2)
\section{Like is friend to like (213d6--215c2)}

Having switched back to Lysis as interlocutor, Socrates proposes that they
follow the wisdom of the poets who teach that ``like befriends like".  In
particular, Socrates quotes a line of verse (poet unnamed) that says ``A
god always leads like to like" (214a6).  (Socrates also notes that the
natural philosophers approve this idea also.)

Socrates immediately picks at the thesis, saying it's at least half false.
The false half concerns bad people.  In the case of bad people, Socrates
argues, they are not each other's friends.  He argues as follows\footnote
{After this argument, Socrates adds a characteristic flourish: Bad people
aren't really like anyone, not even themselves (214c7-d1).  Therefore, when
the poets say like is to like, they are riddling and really only mean
\word{good} like people.}:

\begin{enumerate}

    \item A bad person does wrong [to other people] (214c2).

    \item People doing wrong and the people they do wrong to cannot be
        friends (214c2--3).

\end{enumerate}

Following this argument to dismiss bad people, Socrates says that he still
has some trouble with the argument.  He argues thus:

\begin{enumerate}

    \item If two people are like each other, then each cannot harm or
        benefit the other in any new way.  The one could already do it
        alone (214e3--215a1).

    \item But if this is true, then there's no cause for affection because
        affection follows benefit (215a1--2).

    \item Where there is no affection, there is no friendship (215a3).

    \item Therefore, like is not friend to like.

\end{enumerate}

Socrates then considers whether good people are friends not insofar as they
are similar to each other but insofar as they are good.  However, he
rejects this possibility with an argument much like the one above.  Good
people are ``capable"\footnote{His word is {\g ἱκανός}.}  And insofar as
someone is capable, that person does not need anything.  No need, no
affection.  The rest as above, more or less (215a6--b8).  The one new idea
is that a friendship between good people could never get started: When they
are apart, they don't desire one another, and when they are together, they
have no need of one another.  The importance of where the chain of desire
\emph{begins} will return later in the dialogue.

% }}} Like is friend to like (213d6--215c2)

% {{{ Unlike is friend to unlike (215c3--216b9)
\section{Unlike is friend to unlike (215c3--216b9)}

Socrates regroups again, and again he turns to a poet.  He remembers that
he once heard someone say that similar people and good people were natural
enemies to each other.  That is, like quarrels with like, and a good person
is an enemy to another good person.\footnote{I feel reasonably sure that
the second part of this marks the following argument as potentially
unSocratic.  But see below.} This person used the Hesiodic wisdom that
``potter hates potter" (\book{Works and days 25}).  The idea is that
similarity brings out rivalry and hatred while different people fit each
other's needs.  So a weak person needs a strong one for protection, and
a sick person needs a doctor.  Socrates also cites some common-sense
notions that dry needs wet, cold needs hot and so on.  Menexenus agrees
that this sounds good, and they agree to investigate the thesis that unlike
is friend to unlike.

But Socrates immediately worries that eristic debaters will attack and make
fools of them.  The argument he presents in their voice quickly squashes
this new answer to the question.  These eristics, he imagines, will say
that enmity is most unlike friendship.  They will then ask ``Is enemity
friends with friendship, or friendship friendly with enmity?"  Similarly is
justice friends with injustice or wisdom with foolishness or good with
evil?  These are all absurd conclusions, and so Menexenus and Socrates
abandon the claim that unlike is friend to unlike.

% }}} Unlike is friend to unlike (215c3--216b9)

% {{{ What is neither good nor bad (216c1-218c3)
\section{What is neither good nor bad (216c1-218c3)}

Socrates tries to break through this problem by proposing that what becomes
a friend of the good is something that is ``neither good nor bad"
(216c1--2).  Menexenus doesn't follow, and Socrates admits that he is
himself somewhat loopy from all the different possibilities they've
considered.  Nevertheless, I think he's serious here.

Socrates argues thus initially:

\begin{enumerate}

    \item There are three categories: good, bad and neither (216d5--7).

    \item As already shown, good is not friend to good, nor bad to bad nor
        good to bad (216d7--e1).

    \item The remaining options are (1) neither good nor bad is friend to
        the good or (2) to something else neither good nor bad.  (It is
        summarily ruled out that it might be friends to what is bad.)
        (216e1--4).

    \item But a previous argument just showed that like is not friend to
        like.  Therefore, what is neither good nor bad is friend to what is
        good (216e5--217a2).

\end{enumerate}

In order to see whether this argument holds up, Socrates considers the case
of health and the body.  In and of itself, Socrates says, a body is neither
good nor bad.  A healthy body does not need medical care, and no one who is
healthy befriends a doctor because of his health.  However a sick person
does befriend a doctor because of his illness.  So, Socrates generalizes,
that which is neither good nor bad befriends what is good as a result of
the presence of something bad.

Socrates makes the further point that the thing which is neither good nor
bad can become bad---or bad enough?---that it no longer desires the good.
In order to make more sense of this, Socrates considers hair
dye.\footnote{Yes, you read that right.}  Some things can be present in
something without making that something truly as they are.  For example,
your hair can have dye in it without truly becoming the color of the dye.
So in the case of something neither good nor bad.  It can have some bad in
it without truly becoming bad.  But if it does become truly bad, then it
will no longer desire or befriend the good.

Socrates describes philosophy using this same analysis.  The wise don't do
philosophy, presumably because they don't need it.  Nor do people who are
so ignorant that they have become bad.  The people who do philosophy are
bad insofar as they are ignorant, but they still believe that they do not
know the things that they do not know.  Parts of this sound very familiar.
In particular the people who believe that they don't know what they don't
know sound very much like Socrates as he describes himself in
\book{Apology}.  But I'm not sure how to understand the idea that ``good
people don't philosophize" (218a2-4 and 218b3) as a serious Socratic
proposal.

% }}} What is neither good nor bad (216c1-218c3)

% {{{ An infinite regress? (218c4--220e6)
\section{An infinite regress? (218c4--220e6)}

Socrates rejoices briefly at the success of the previous argument, but he
immediately has a concern.  He worries that the argument has unfortunate
implications.  As he explains:

\begin{enumerate}

    \item Whatever is a friend, is a friend (1) to something, (2) for the
        sake of something and (3) because of something.  The difference
        between (2) and (3) is initially confusing, but Socrates explains
        with the example of illness.  A person is a friend to a doctor (2)
        for the sake of health, but (3) because of illness.  The illness
        causes him to want health, and thus indirectly to want the doctor's
        help (218d6--219a4).

    \item Using the terminology from the previous discussion, Socrates uses
        this structure to describe friendship: ``that which is neither good
        nor bad is (1) a friend of the good (3) because of the bad (2) for
        the sake of the good and friendly" (219a3--4).  He then
        reformulates this further as ``The friend is a friend of the friend
        for the sake of the friend because of the enemy" (219b2--3).

    \item The wording of the last version suggests a number of problems.
        First, is runs afoul of the earlier argument that like is not
        friend to like, but Socrates dismisses this fear for the moment.
        Socrates worries more about the implications of the ``for the sake
        of" clause.  If the thing we are pursuing is also a friend, then
        \emph{it} must be sought in turn for the sake of something and an
        infinite regress threatens.  So Socrates wonders if there is
        something that can serve as a beginning of the chain without itself
        referring to anything else.  Socrates also says that all of the
        other things are ``friends" only by a convenient figure of speech.
        The only thing that is truly a friend is the one that grounds all
        the others (219d2--5).

\end{enumerate}

Socrates expands on the final point with two examples.  First example: A father
loves his son.  As a result, he may love many other things which he believes
can help his son.  But he does not really love those other things \foreign{per
se}. He loves them only insofar as they contribute to his son's wellbeing.
Second example: people love money (Socrates actually says ``gold and silver").
But really what people love are the things that money can get.

By applying this argument to friendship, Socrates solves one problem.  Since
the objects other than the primary one are not really loved (or ``beloved",
that is {\g φίλος}), then we can escape the fear that ``friend is friend of
friend on behalf of friend".  This means that we are safe, I suppose, from the
earlier arguments that ``like is not friend to like".

As a next issue, Socrates takes back up the good.  If the good is a friend ({\g
φίλον}), is that because of the bad?  As Socrates argues, the answer is yes.
Only because of something bad is the good useful or necessary.  The good is
like a drug: Without illness it would be useless.

Therefore, the friend at the start of the chain is unlike all the other things
we call friends.  The first friend is only a friend because of bad and harm,
but all the others are friends because of some other friend and benefit.
I believe that saves us both from the infinite regress problem and the problem
that Socrates quickly passed over above.\footnote{I suspect Plato set things up
that way deliberately.  Initially Socrates says ``Let's not worry about this
part." and then later Socrates solves that problem.  Thus, he was justified in
ignoring the problem \foreign{ex post facto}.}

% }}} An infinite regress? (218c4--220e6)

% {{{ Desire is the cause of friendship (220e6--221d6)
\section{Desire is the cause of friendship (220e6--221d6)}

Socrates immediately takes aim at the claim that {\g κακόν} (bad, harm) is
the cause of friendship.  He argues as follows:

\begin{enumerate}

    \item Sometimes desires stem from harm, but not always.  Therefore even
        if harms disappear, some desires will remain (220e6--221b6).

    \item Whoever desires and loves something is friendly towards that
        something (221b7--8).

    \item Therefore, even if all harm disappears, there can still be some
        friends (221b8--c1).

    \item If harm were the cause of friendship, this would be impossible
        since something cannot occur if its cause disappears (221c2--5).

    \item Therefore, they were wrong to believe that harm is the cause of
        friendship, since friendship can survive the (imagined)
        disappearance of harm (221c5--d2).

    \item Desire is the true cause of friendship (221d2--6).

\end{enumerate}

This argument has at least two significant problems.  First, in the first step,
Socrates seems to waffle between a desire being for something harmful and the
desire itself being harmful.  I'm not sure if this is a culpable ambiguity or
not since I can't entirely follow the first step. (I've paraphrased it above to
make it more comprehensible.)  Second, as \citet{watt1987} points out, the
final step is a non-sequitur.  I can't see why Socrates thinks that he's
entitled to it.  My best guess is something like this:

\begin{enumerate}

    \item There must be a univocal cause of friendship (axiomatic?).

    \item The argument above proves that there is sometimes friendship where
        there is not harm and that in those cases desire is the cause of the
        friendship.

    \item Therefore, desire must always be the cause of friendship and not
        harm.

\end{enumerate}

It might be worthwhile to go back and try to reanalyze the cases from the
previous argument, replacing \word{harm} with \word{desire}.  (For example, how
would the case of the sick person look?  The sick person is a friend of
a doctor for the sake of health because of his desire to be well?  In that
case, the \phrase{for the sake of} and \phrase{because of} clauses seem to
collapse into one another.)

% }}} Desire is the cause of friendship (220e6--221d6)

% {{{ Genuine love is necessarily reciprocated (221d6--222b2)
\section{Genuine love is necessarily reciprocated (221d6--222b2)}

Socrates next argues towards the conclusion that ``It is necessary that the
legitimate and real lover to be loved by his beloved" (222a6--7).  The
argument proceeds as follows.  The first step:

\begin{enumerate}

    \item Desire is for that which one needs (221d6--e1).

    \item Something in need is, therefore, a friend of what it needs
        (221e1--2).  Friendship follows desire, according to the previous
        argument.

    \item Need comes about when something has been taken away (221e2--3).
        This is new, and I'm not sure how persuaded we should be by it.

    \item Love, friendship and desire, therefore, are for what is one's own
        (221e3--5).  (Because one desires what one needs, and need comes to
        be when something has been taken from one.  The thing taken was
        presumably ``one's own" before it was taken?)

\end{enumerate}

Once the first step is agreed to, Socrates continues:

\begin{enumerate}

    \item If someone desires, loves or befriends another, it must be
        because the other is like the lover in some way
        (221e7--222a3).\footnote{This may be odd: Menexenus agrees to this
        claim immediately and emphatically, but Lysis is silent.  I suspect
        this is a small joke: Lysis is in love with Menexenus, and he
        suspects where the argument is going?  In any case, his silence does
        seem to matter, thought I'm not completely sure how.}

    \item Reminder that it has been shown that it is necessary to love what
        is one's own (222a5--6).

    \item Therefore, it is necessary for the true and not feigned lover to
        be loved by his beloved (222a6--7).  If person A truly loves person
        B, that must be because person B is like person A in some sense.
        If they are like, then they share what is one's own?  Therefore,
        the likeness and the necessity of loving would run both ways.
        Hence, true love must be reciprocated.

\end{enumerate}

Both Lysis and Menexenus agree to this conclusion only grudgingly.
Hippothales, however, seems delighted by it.

% }}} Genuine love is necessarily reciprocated (221d6--222b2)

% {{{ Is one's own the same as like? (222b3--d8)
\section{Is one's own the same as like? (222b3--d8)}

Socrates wonders now whether ``What is one's own differs from the like?"
This is a concern since they already proved that like is not friend to like
(see notes on 213d6--215c2 above).

The argument runs thus:

\begin{enumerate}

    \item First, Socrates gets the boys to agree that the two concepts are
        different (222c1--3).

    \item Socrates then asks ``Is good one's own for everything and evil
        other to everything? Or are good, bad and the neither good nor bad
        each own's own to itself?"  The boys answer that each is one's one
        for itself (222c3--d1).

    \item But this leads to a very bad result: The unjust will be friends
        with the unjust no less than the good with the good.  As Socrates
        reminds the boys, they shrank from this conclusion earlier, so it
        should trouble them now (222d1--5).

    \item Another bad result: If they say that the good and one's own are
        the same, then the good man will be friends only with a good man.
        But this too they disproved earlier (222)

\end{enumerate}

% }}} Is one's own the same as like? (222b3--d8)

% {{{ Aporetic conclusion (222e1--223b8)
\section{Aporetic conclusion (222e1--223b8)}

Much like \book{Euthyphro}, Socrates proposes a thorough review of the
case, but people immediately have to leave.  Socrates quickly runs through
many of the options they have exhausted and he expresses his own complete
confusion.  His next thought is to involve someone older in the discussion,
but at that point the boys' slaves remind them that it is time to go home.
(The slaves are a bit drunk from the festival, so it's impossible to
convince them to stay.  They force their opinion on the group, which
perhaps recalls the initial idea that Lysis is ruled by slaves? An odd
detail.)

Socrates ends with the famous lines:

\begin{quote}

    {\g Νῦν μέν, ἦν δ᾽ἐγώ, ὦ Λύσι τε καὶ Μενέξενε, καταγέλαστοι γεγόναμεν
        ἐγώ τε, γέρων ἀνήρ, καὶ ὑμεῖς.  ἐροῦσι γὰρ οἵδε ἀπιόντες ὡς οἰόμεθα
        ἡμεῖς ἀλλήλων φίλοι εἶναι---καὶ ἐμὲ γὰρ ἐν ὑμῖν τίθημι---οὔπω δὲ
        ὅτι ἔστιν ὁ φίλος οἷοί τε ἐγενόμεθα ἐξευρεῖν} (223b4--8)

    Now, I said, Lysis and Menexenus, we have proven to be ridiculous, both
    I, an old man, and you.  For these men as they depart will say that we
    think that we are friends of each other---for I place myself among
    you---but that we have not yet been able to discover what a friend is.

\end{quote}

People frequently cite this as an instance of the (alleged) Socratic belief
that you cannot know that x is F if you do not know what it is to be F.
This belief is called the \phrase{priority of definition} or something
similar throughout the literature.  \citet{geach1966} calls it \phrase{the
Socratic fallacy}.  The conclusion to \book{Lysis} plays a central role in
what we think about Socrates and this principle or fallacy.

% }}} Aporetic conclusion (222e1--223b8)


\newpage
\bibliographystyle{apa}
\bibliography{plato}

\end{document}
% }}}
