% [[- LaTeX prelude
\documentclass[12pt,letterpaper]{article}

\usepackage[no-math]{fontspec}
\setmainfont{Baskerville}

\usepackage[nolocalmarks]{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage[variant=ancient]{latin}
\setotherlanguage[variant=ancient]{greek}
\newfontfamily\greekfont[Script=Greek]{Times New Roman}

\usepackage{csquotes}
\usepackage[style=authoryear,backend=biber,dashed=false,%	
	abbreviate=false,maxnames=4]{biblatex}
\addbibresource{plato.bib}
% See the following resources:
%   https://orthostat.wordpress.com/2013/02/16/moving-to-biblatex-1/
%%
% Remove pp. in citations.
\DeclareFieldFormat{postnote}{#1}
% Remove pp. from bibliography.
\DeclareFieldFormat%
    [article,inbook,incollection,inproceedings,thesis,patent,unpublished]%
    {pages}{{\nopp#1}}
% Use single quotes around titles of chapters and articles, not double.
% \DeclareFieldFormat%
%     [article,inbook,incollection,inproceedings,thesis,patent,unpublished]%
%     {title}{`#1'}
% Use commas rather than periods to separate units in the bibliography.
% \renewcommand*{\newunitpunct}{\addcomma\space}
% Place a space rather than a comma between name and year in citations.
% \renewcommand{\nameyeardelim}{~}
% Place a colon+narrow space before page numbers in citations.
\renewcommand{\postnotedelim}{:\,}
% No "in:" for articles at all. For all else, remove the colon.
\renewbibmacro{in:}{%
	\ifentrytype{article}{}{\printtext{\bibstring{in}}}
}
% Remove parentheses around date in bibliography.
% See http://tex.stackexchange.com/a/40710/29387
\usepackage{xpatch}
\xpatchbibmacro{date+extrayear}{%
	\printtext[parens]%
}{%
	\setunit{\addperiod\space}%
	\printtext%
}{}{}

\usepackage{enumitem}
\setlist{noitemsep}
\usepackage[super]{nth}


\begin{hyphenrules}{latin}
    \hyphenation{}
\end{hyphenrules}

\begin{hyphenrules}{greek}
    \hyphenation{ἐκ-μανθάν-ειν}
\end{hyphenrules}

%\pagestyle{plain}
% -]] Latex prelude

% [[- LaTeX document
\begin{document}

% [[- Title page
\begin{titlepage}
\title{Notes on Plato's \textit{Phaedo}}
\author{Peter Aronoff}
\date{May 2017}
\maketitle
\thispagestyle{empty}
\end{titlepage}
% -]]

% [[- Characters and setting
\section{Characters and setting}

% [[- Characters
\subsection{Characters}

\subsubsection{Frame}

The frame narrative contains two speakers, Echecrates and Phaedo. Echecrates is from Phlius in the Pelopponesus; he is also a Pythagorean although the dialogue never explicitly mentions this. We don't learn much about him, but he is clearly interested in Socrates even if he isn't a formal follower of his. Phaedo is a Socratic. He is from Elis, and apparently was a prisoner of war at Athens during the Pelopponesian war, but he was freed with the help of Socrates.\footcite[7]{rowe1993} \textcite[231]{nails2002} argues that Phaedo should be about 20 at the dramatic time of the dialogue. He appears to be making his way home to Elis from Athens when he meets Echecrates in Phlius.

\subsubsection{Death scene}

We briefly see Xanthippe, Socrates's wife, and Crito, his good friend, but the main characters are Socrates, Simmias and Cebes. Simmias and Cebes, both originally from Thebes, are somehow associated with (\textgreek{συγγίγνεσθαι} 61d) Philolaus, a Pythogorean, but Rowe argues that the two men themselves are not necessarily Pythagoreans \parencite[7]{rowe1993}. He says ``associate with'' is vague, and he also argues that Simmias advances a theory that conflicts with Pythagoreanism, namely the harmony theory of the soul. However, Rowe also admits that Echecrates, who certainly was a Pythagorean, is all in favor of the theory. So I'm not sure what support of that theory shows. (Also note that \textcite[138]{nails2002} considers Echecrates as also a ``student of Philolaus.'') Both Simmias and Cebes are relatively young (\textgreek{νεανίσκος} 89a) at the time of Socrates's death, but otherwise, we don't know anything about when they lived.

% -]] Characters

% [[- Setting
\subsection{Setting}

\subsubsection{Frame}

The speakers are in Phlius, and Phaedo appears to be heading from Athens to Elis after the death of Socrates. We have no specific information about where in Phlius the conversation takes place.

\subsubsection{Death scene}

It is the day of Socrates's death in 399, and the speakers meet at the jail where Socrates was held after his trial.

% -]] Setting

% -]] Characters and setting

% [[- Outer frame, part 1 (57--59c7)
\section{Outer frame, part 1 (57--59c7)}

Echecrates begins the dialogue by asking whether Phaedo was there when Socrates died or whether he heard the story from someone else. This is similar to other dialogues with frames. Listeners often hope to get their stories as directly as possible, and if the storyteller wasn't personally present, they will often give a more or less detailed explanation of how they got their information. In this way, the dialogues enact a kind of submission of \textit{bona fides}.

In addition, such introductions often allow Plato to slip in other material that helps to guide us as readers. In this case, Phaedo says that he was there himself, but he also tells us (1) why there was such a long delay between the trial and execution and (2) how strange an experience the day of Socrates's death was.

An annual religious duty caused the delay. Every year the Athenians honor Apollo by sending a ship and offering to Delos; they view this as carrying out an eternal obligation to Apollo for saving the life of Theseus when he traveled to Crete. During the time that the ship is away, the city is in a state of heightened purification, and so they cannot perform executions until the ship returns. Some years this takes a while since the ship can be stuck at Delos waiting for the winds to be right, and that is what happens in this case. This is the first way in which Apollo features in the dialogue.

Phaedo experiences an emotional state that he finds strange (\textgreek{θαυμάσια} 58e1) on the day of Socrates's execution. On the one hand, he doesn't feel pity, as might have been expected at the death of a close friend, because Socrates himself appeared to be so happy. But by the same token, he didn't enjoy the philosophical conversations of the day, as he normally would have, since the imminent death of Socrates overshadowed everything. As a result, Phaedo (and many of the others present) experienced an ``entirely unusual emotion'' (\textgreek{ἀτεχνῶς ἄτοπόν τι\dots πάθος (59a5)}), a blending of of pleasure and pain.\footnote{This seems to anticipate when Socrates says that pleasure and pain have a kind of connected relationship at the start of the narrative (60b ff.), but the exact connection between the ideas is unclear. Phaedo and Socrates seem to disagree, though commentators have been quick to say that there's no formal contradiction. See \cite[\textit{ad} 59a1--b4 and 60b1--c7]{gallop1975} and \cite[\textit{ad} 59a4, 60b4, and 60b5]{rowe1993}.}
% -]] Outer frame, part 1 (57--59c7)

% [[- Two attitudes towards death (59c8--69e7)
\section{Two attitudes towards death (59c8--69e7)}

% [[- Pleasure and pain (60b1--c7)

\subsection{Pleasure and pain (60b1--c7)}

After a rapid narration of the morning meeting of Socrates's friends outside the jail and the quick dismissal of Xanthippe, Phaedo focuses on Socrates immediately after he has been unchained.\footnote{He is unchained in preparation for his execution. It is worth saying, however, that he doesn't appear to be chained in, e.g., \textit{Crito}. I wonder if the detail about the chains isn't merely a way for Plato to drop in this weird discussion of pleasure and pain.} Phaedo describes in surprising detail how Socrates is sitting, and he then recounts a strange observation Socrates makes about pleasure and pain. I've never been able to get a handle on what the sitting means,\footnote{Though note that there's more attention to how Socrates sits in the near sequel to this passage at 61c10--d2. I don't know what that means either.} but I want to look at least briefly at the pleasure and pain bit.

After the jailers have released Socrates from his shackles, he rubs at his legs, and he remarks that ``what people call pleasure'' is strange (\textgreek{ἄτοπον} 60b3). In particular, this alleged pleasure is quite odd when we consider its apparent opposite, pain. The two of these things refuse to exist in one person at the same time, and yet if a person experiences one, then the other is bound to follow. If Aesop had noticed this phenomenon, he could have told a story about how a god wished to end their battling, but settled for joining them together by the heads. If this were the end of it, we would probably take this as just a small bit of versimilitude in Plato's narration. But this isn't the end of it.

Opposites and their relation to one another come up repeatedly in the dialogue, both before and after this section. Phaedo has already discussed his mixed emotional state (58d--59a). And later Socrates will rely on our intuitions about opposites when he argues that life and death form an eternal cycle (69e-72d). When Socrates explains the role of forms in causation (99d4 ff.), he will again make use of intuitions about opposites: whether, for example, something could be larger by the presence in it of smallness.

So, in a nutshell, although I have no overall view of \textit{what} Plato is doing in this dialogue with opposites, I'm convinced that critics have not paid it enough attention. In particular, readers are in a hurry to reconcile what Phaedo says about opposed emotions with what Socrates says about pleasure and pain. They are, however, narrowly concerned to demonstrate that the two men don't contradict each other. As a result, they don't devote sufficient attention to what we should do with the plurality of references to the relationship of opposites.
% -]] Pleasure and pain (60b1--c7)

% [[- Evenus's question (60b8--61c5)
\subsubsection{Evenus's question (60b8--61c5)}

At the mention of Aesop, Cebes remembers a question that Evenus\footnote{Evenus is mentioned as a well-known rhetorician in \textit{Apology} (20b) and \textit{Phaedrus} (267a).} had asked him: why has Socrates recently turned to writing poetry? Apparently Socrates had versified some of Aesop's fables, and he also wrote a (prefatory? \textgreek{προοίμιον} 60d2) poem to Apollo. Evenus and others asked Cebes about this. It seems that all these people find it surprising for Socrates to have done this, but it is not clear whether that is because he wrote verse or because he wrote at all.

Socrates replies that he has been writing in response to dreams that he takes to be messages from the gods. In these dreams he is told to make poetry. Earlier in his life, he interpreted ``poetry'' to mean philosophy, and he thought that the repetition of the dream---and the injunction---was similar to bystanders cheering on runners. When the bystanders yell ``Run, run!'' they aren't telling the runners to do something new or different. They are telling the runners to continue doing what they are already doing. But in order to be as cautious as possible, Socrates has recently decided to take the dreams more literally and to write poetry. He wrote to Apollo since a festival for Apollo was happening while he was in jail and then he set some of Aesop's fables into verse since he had no stories of his own to tell.

This short conversation does several things within the dialogue. First, we have an early and clear display of Socratic piety, as well as another example of the equanimity with which Socrates faced his execution. Most importantly, however, it tees up Socrates to offer Evenus some parting wisdom. Socrates tells Evenus that he should follow Socrates, i.e., that he should die, as quickly as possible. When Cebes is shocked at this advice, Socrates says, more or less, ``Well, Evenus is a philosopher, right?'' For Socrates, it will turn out, if you are a philosopher, it is natural for you to want to die. This will form the direct and indirect subject of the next few sections of the dialogue.
% -]] Evenus's question (60b8--61c5)

% [[- Philosophers, death, and suicide (61c6--?)
\subsubsection{Philosophers, death, and suicide (61c6--?)}

Socrates says that if Evenus or anyone else is truly a philosopher, they will be willing to die although they won't commit suicide. Cebes asks Socrates to elaborate; it is not initially clear whether Cebes is more surprised at the ban on suicide, which he mentions first, or on the conjunction of the ban on suicide plus the philosopher's willingness to die. In any case, as they begin to discuss the matter, Cebes turns first to the ban on suicide, and so that is what Socrates replies to first.

Unfortunately Socrates begins his answer with what \textcite[\textit{ad} 62a2--7]{gallop1975} calls an ``intractable'' and ``much disputed'' sentence. In a four-and-a-half page note, he runs through five interpretations of the Greek. Gallop himself inclines towards the last, and the key points of his interpretation are these:

\begin{enumerate}
    \item There are two main items up for consideration, which Gallop calls (P) and (Q). He subdivides (P) into three pieces since different interpretations take different views about the attitudes of Socrates and Cebes towards these sub-parts: (P1) whether suicide alone of all things is unqualified, (P2) whether it never happens, as it does for other things on some occcasions and for some people (P3) that it is better for a person to be dead than alive, and (Q) those people for whom it is better to be dead nevertheless may not benefit themselves by killing themselves; instead they must await ``another benefactor.''
    \item \textcite[82]{gallop1975} claims ``Good sense is obtained by reading the sentence purely as an expression of the doubts in Cebes' mind''.
    \item In addition, Socrates is not necessarily committed to the the truth of (P) or (Q).
\end{enumerate}
% -]] Philosophers, death, and suicide (61c6--?)

% -]] Two attitudes towards death (59c8--69e7)

\newpage
\printbibliography

\end{document}
% -]]
